%!TEX root = forallx.tex
\thispagestyle{empty}
{\Huge\forallx}

{\tt An Introduction to Formal Logic}

\vfill


{\sf P.D. Magnus}\\
\emph{University at Albany, State University of New York}


\vfill


{\sf
	\href{https://www.fecundity.com/logic/}{fecundity.com/logic}, version 1.4 [\bookversion]\\
	This book is offered under a Creative Commons license.\\
	(Attribution 4.0)
}



\newpage
\thispagestyle{empty}%
{\sf
The author would like to thank the people who made this project possible. Notable among these are Cristyn Magnus, who read many early drafts; Aaron Schiller, who was an early adopter and provided considerable, helpful feedback; {and} Bin Kang, Craig Erb, Nathan Carter, Wes McMichael, Selva Samuel,  Dave Krueger, Brandon Lee, Toan Tran, Marcus Adams, Matthew Brown, and the students of Introduction to Logic, who detected various errors in previous versions of the book.
}

\vfill
{
\copyright\ \ifthenelse{\year=2005}{\number\year}{2005--\number\year} by P.D. Magnus. Some rights reserved.
}

{\footnotesize
You are free to copy this book, to distribute it, to display it, and to make derivative works, under the following condition: Attribution. You must give the original author credit. --- For any reuse or distribution, you must make clear to others the license terms of this work. Any of these conditions can be waived if you get permission from the copyright holder. Your fair use and other rights are in no way affected by the above. --- This is a human-readable summary of the full license, which is available on-line at \url{http://creativecommons.org/licenses/by/4.0/}
}

{
Typesetting was carried out entirely in \LaTeX$2\varepsilon$. The style for typesetting proofs is based on fitch.sty (v0.4) by Peter Selinger, University of Ottawa.

This copy of \forallx\ is current as of \today. The most recent version is available on-line at \url{http://www.fecundity.com/logic}
}

{
A magyar változatot a Pécsi Tudományegyetem Természettudományi Karának hallgatói és oktatói készítették.

Fordították:
Andrikó Lívia, Arany Enikő,
Bakos Marcell, Bali Gina, Balázs Bálint Balázs, Bank Bence Boldizsár, Bárány Miklós,
Békefi Benjámin, Bolboaca Sergiu, Böndicz Anna, Bruckman Denton, Buzás Dániel,
Cári Zoltán Gergő, Csanády Bettina, Császár Szabolcs, Csipszer András, Czigány Elliot,
Dankó Dominik, Dávid Gergő, Dávid Rebeka, Erdélyi Tamás Gergely,
Farkas Nikolett, Fövenyessy Vivien,
Gajda Réka, Gál Zalán, Geresdi Bence, Gömöri Marin, Gulyás Fanni, Gyovai-Rafa Jeremy, Győr Richárd,
Harsányi Magor, Hegedűs Ákos, Herger Dávid, Hetesi Dorina, Hollósi Tamás, Horváth András István, Horváth Patrik,
Jobb Tamás, Juhász Krisztián,
Kertész Marcell Loránd, Kiskovács Barna, Kiss Martina Anita, Klopfer Máté, Kocziány Tamás, Kollár Márk, Kovács Norbert, Kozma Olivér, Kövecses Ferenc, Krepler Péter, Kun Máté,
László Csenge Zorka, Lovász Levente Márk, Lőrinczy Dorottya Rebeka,
Magyar Arnold, Martin Dávid, Marton Balázs, Máté Dávid, Máté Patrik, Mischl-Ruzsik Barbara, Molnár Bence Miklós, Molnár Karolin, Molnár Rajmund, Muzsek Balázs,
Nagy Attila Szabolcs, Nagy Gábor, Nagy Veronika, Nemes Csaba, Németh Gábor, Németh Máté, Nichter Valter, Novák Nikolett,
Orbán Mia, Örömi Dániel,
Páj Gábor, Pénzeli Gabriella, Perlaki Balázs, Pethő Eszter, Pintér Regina, Pongrácz Ádám,
Rózsa Aliz, Rózsavölgyi Kitti,
Sáfrány Regő, Sarek Gergő, Sárosi Dávid, Schäffer Gergő, Schmidt Péter Henrik, Spán Richárd, Sudár Tamás,
Szarka Veronika, Szijártó György Ágoston,
Tolnai Balázs, Tőzsér Bence Alex, Turjányi Kolos Kálmán, Ujvári Dávid, Urbán Emese,
Varga Bálint, Varga Márton, Vass Vilmos, Wágner Norbert, Zakariás Bálint, Zouaoui Leila, Zsók Győző.
}